---
title: "Example of Methylome Analysis with MethylIT using Cancer Datasets"
subtitle: <h1>A [Methyl-IT](https://github.com/genomaths/MethylIT) user guide. 
           A guide for users</h>
author: |
 | Robersy Sanchez$^1$
 | rus547@psu.edu
 |
 | Sally Mackenzie$^1$
 | sam795@psu.edu
 |
 | $^1$Department of Biology and Plant Science. 
 | Pennsylvania State University, University Park, PA 16802
date: "`r format(Sys.time(), '%d %B %Y')`"
fontsize: 11pt
fontfamily: "serif"
bibliography: bibliography.bib
output:
  rmarkdown::html_document: 
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
      smooth_scroll: true
    number_sections: true
    highlight: tango
    theme: united
    geometry: margin=1in
    keep_md: yes
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  
abstract: |
  Methyl-IT, a novel methylome analysis procedure based on information
  thermodynamics and signal detection was recently released. Methylation
  analysis involves a signal detection problem, and the method was designed to
  discriminate methylation regulatory signal from background noise induced by
  thermal fluctuations. Methyl-IT enhances the resolution of genome methylation
  behavior to reveal network-associated responses, offering resolution of gene
  pathway influences not attainable with previous methods. Herein, an example
  of MethylIT application to the analysis of breast cancer methylomes is
  presented.

---


<style type="text/css">

.list-group-item.active, .list-group-item.active:focus, 
.list-group-item.active:hover {
    z-index: 2;
    color: #fff;
    background-color: #337ab7;
    border-color: #337ab7;
}

.tocify-subheader .tocify-item {
  font-size: 0.80em;
  padding-left: 25px;
  text-indent: 0;
}

.tocify-header .tocify-item {
  font-size: 0.85em;
  padding-left: 20px;
  text-indent: 0;
}

body{ /* Normal  */
      font-size: 18px;
      font-family: "Times New Roman", Times, serif;
      text-align: justify
  }
td {  /* Table  */
  font-size: 8px;
}

h1.title {
  font-size: 38px;
  font-family: "Times New Roman", Times, serif;
  color: DarkRed;
  .subTitle {
  font-size: 24px;
  font-family: "Times New Roman", Times, serif;
  color: DarkRed;
  }
}

h1 { /* Header 1 */
  font-size: 28px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 22px;
    color: DarkBlue;
    font-family: "Times New Roman", Times, serif;
}
h3 { /* Header 3 */
   font-size: 18px;
   color: DarkBlue;
   font-family: "Times New Roman", Times, serif;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(options(width = 80))
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# [Methyl-IT](https://github.com/genomaths/MethylIT)
MethylIT is an R package for methylome analysis based on information
thermodynamics and signal detection. The information thermodynamics-based
approach is postulated to provide greater sensitivity for resolving true signal
from the thermodynamic background within the methylome [@Sanchez2016]. Because
the biological signal created within the dynamic methylome environment
characteristic of plants is not free from background noise, the approach,
designated MethylIT, includes the application of signal detection theory
[@Greiner2000; @Carter2016; @Harpaz2013; @Kruspe2017]. A basic requirement for
the application of signal detection is a probability distribution of the
background noise. Probability distribution, as a Weibull distribution or gamma
distribution models, can be deduced on a statistical mechanical/thermodynamics
basis for DNA methylation induced by thermal fluctuations [@Sanchez2016].
Assuming that this background methylation variation is consistent with a Poisson
process, it can be distinguished from variation associated with methylation
regulatory machinery, which is non-independent for all genomic regions
[@Sanchez2016]. An information-theoretic divergence to express the variation in
methylation induced by background thermal fluctuations will follow a Weibull
distribution model, provided that it is proportional to minimum energy
dissipated per bit of information from methylation change.

Herein, we provide an example of MethylIT application to the analysis of breast
cancer methylomes. Due to the size of human methylome the current example only
covers the analysis of chromosome 13. A full description of MethylIT
application of methylome analysis in plants is given in the manuscript
[@Sanchez2018].

The R packages required in this example are:
```{r libraries, message=FALSE}
library(MethylIT)
library(ggplot2) # graphic
library(reshape2) # To reshape the data frame
library(grid) # For multiple plots
library(gridExtra) # For multiple plots
library(VennDiagram) 
library(rtracklayer) # To import gene annotation
```

# Available datasets and reading
Methylome datasets of whole-genome bisulfite sequencing (WGBS) are available
at Gene Expression Omnibus (GEO DataSets). For the current example, datasets
from breast tissues (normal and cancer) and embryonic stem cells will be
downloaded from GEO. The data set are downloaded providing the GEO accession
numbers for each data set to the function 'getGEOSuppFiles' (see the function
[getGEOSuppFiles](https://genomaths.github.io/MethylIT_HTML_Manual/getGEOSuppFiles.html) 
help).

```{r GEO, include = TRUE, message=FALSE}

# Embryonic stem cells datasets
esc.files = getGEOSuppFiles(GEO = c("GSM2041690", "GSM2041691", "GSM2041692"),
                            verbose = FALSE)
# Breast tissues (normal, cancer, metastasis)
cancer.files = getGEOSuppFiles(GEO = c("GSM1279517", "GSM1279514",
                                       "GSM1279513"), verbose = FALSE)

```

The file path and name of each downloaded dataset is found in the output
variables 'esc.files' and 'cancer.files'.

## _Reading datasets_
Datasets for our example can be read with function [readCounts2GRangesList](https://genomaths.github.io/MethylIT_HTML_Manual/readCounts2GRangesList.html).
To specify the reading of only chromosome 13, we can specify the parameter
'chromosomes = "Chr13"'. The symbol chromosome 13, in this case "Chr13", must
be consistent with the annotation provided in the given GEO dataset. Each file
is wholly read with the setting 'chromosomes = "Chr13"' and then the GRanges
are built only with chromosome 13, which could be time consuming. However,
users working on Linux OS can specify the reading of specific lines from each
file by using regular expressions. For example, if only chromosomes 1 and 3 are
required, then we can set chromosomes = NULL (default) and 'chromosome.pattern
= "^Chr[1,3]"'. This will read all the lines in the downloaded files starting
with the words "Chr1" or "Chr3". If we are interested in chromosomes 1 and 2,
then we can set 'chromosome.pattern = "^Chr[1-2]"'. If all the chromosomes are
required, then set chromosomes = NULL and chromosome.pattern = NULL (default).

```{r reading datset, include = TRUE}
# Embryonic stem cells datasets
ref = readCounts2GRangesList(filenames = esc.files,  
                             sample.id = c("ESC1","ESC2", "ESC3"),
                             columns = c(seqnames = 1, start = 2,
                                         mC = 4, uC = 5), pattern = "^chr13",
                             remove = TRUE, verbose = FALSE)
# Breast tissues (normal, cancer, metastasis)
LR = readCounts2GRangesList(filenames = cancer.files,
                            sample.id = c("Breast_normal", "Breast_cancer",
                                          "Breast_metastasis"),
                            columns = c(seqnames = 1, start = 2,
                                         mC = 3, uC = 4),
                            remove = TRUE, pattern = "^13",
                            chromosome.names = "chr13", verbose = FALSE)
```

```{r, include = FALSE}
file.remove(cancer.files, esc.files)

```

In the metacolumn of the last GRanges object, mC and uC stand for the
methylated and unmethylated read counts, respectively. Notice that option 
'remove = TRUE' remove the decompressed files (default: FALSE, see 
?readCounts2GRangesList for more details about this function).

# The reference individual
Any two objects located in a space can be compared if, and only if, there is a
reference point (a coordinate system) in the space and a metric. Usually, in
our daily 3D experience, our brain automatically sets up the origin of
coordinates equal to zero. The differences found in the comparison depend on
the reference used to perform the measurements and from the metric system. The
space where the objects are located (or the set of objects) together with the
metric is called metric space.

To evaluate the methylation differences between individuals from control and
treatment we introduce a metric in the bidimensional space of methylation
levels $P_i = (p_i, 1-p_i)$. Vectors $P_i$ provide a measurement of the
uncertainty of methylation levels. However, to perform the comparison between
the uncertainty of methylation levels from each group of individuals, control
$(c)$ and treatment $(t)$, we should estimate the uncertainty variation with
respect to the same individual reference on the mentioned metric space. The
reason to measure the uncertainty variation with respect to the same reference
resides in that even sibling individuals follow an independent ontogenetic
development. This a consequence of the "omnipresent" action of the second law
of thermodynamics in living organisms. In the current example, we will create
the reference individual by pooling the methylation counts from the embryonic
stem cells.

It should be noticed that the results are sensitive to the reference used. The
statistics mean, median, or sum of the read counts at each cytosine site of
some control samples can be used to create a virtual reference sample. A virtual 
reference individual can be built with function [poolFromGRlist](https://genomaths.github.io/MethylIT_HTML_Manual/poolFromGRlist.html).
It is up to the user whether to apply the 'row sum', 'row mean' or 'row median' of
methylated and unmethylated read counts at each cytosine site across
individuals. Notice that when "*mean*" is selected the virtual reference individual
is the group centroid, which play a fundamental role in multivariate statistic.

```{r pooling, include = TRUE}
Ref = poolFromGRlist(ref, stat = "mean", num.cores = 6L, verbose = FALSE)

Ref

```

Only direct lab experiments can reveal whether differences detected with
distinct references outside the experimental conditions for control and
treatment groups are real. The best reference would be estimated using a subset
of individuals from control group. Such a reference will contibute to remove
the intragroup variation, in control and in treatment groups, induced by
enviromental changes external to or not contralled by the experimental
conditions.

Methylation analysis for each cystosine position is frequently performed in the
bidimensional space of $(methylated, unmethylated)$ read counts. Frequently,
Fisher test is applied to a single cytosine position, under the null hypothesis
that the proportions
$p_{ct}=methylated_{ct}/(methylated_{ct}+unmethylated_{ct})$ and
$p_{tt}=methylated_{tt}/(methylated_{tt}+unmethylated_{tt})$ are the same for
control and treatment, respectively. In this case, the implicit reference point
for the counts at every cytosine positions is 
$(methylated = 0,unmethylated = 0)$, which corresponds to the point 
$P_i = (0, 1)$.

In our case, the Hellinger divergence (the metric used, here) of each
individual in respect to the reference is the variable to test in place of
$(methylated, unmethylated)$ read counts or the methylation levels
$P_i = (p_i, 1-p_i)$.

The use of references is restricted by the thermodynamics basis of the the
theory. The current information-thermodynamics based approach is supported on
the following postulate:

  *"High changes of Hellinger divergences are less frequent than low changes,
  provided that the divergence is proportional to the amount of energy required
  to process one bit of information in methylation system"*.

The last postulate acknowledges the action of the second law of thermodynamics
on the biomolecular methylation system. For the methylation system, it implies
that the frequencies of the information divergences between methylation levels
must be proportional to a Boltzmann factor (see supplementary information from
reference [@Sanchez2016]). In other words, the frequencies of information
divergences values should follow a trend proportional to an exponential decay.
If we do not observe such a behaviour, then either the reference is too far
from experimental condition or we are dealing with an extreme situation where
the methylation machinery in the cell is disfunctional. The last situation is
found, for example, in the silencing mutation at the gene of
cytosine-DNA-methyltransferase in Arabidopsis *thaliana*. Methylation of
5-methylcytosine at CpG dinucleotides is maintained by MET1 in plants.

In our current example, the embryonic stem cells reference is far from the
breast tissue samples and this could affect the nonlinear fit to a Weibull
distribution (see below). To illustrate the effect of the reference on the
analysis, a new reference will be built by setting:

```{r new reference}
Ref0 = Ref
Ref0$uC <- 0

```
The reason for the above replacement is that natural methylation changes
(Ref$mC) obey the second law of thermodynamics, and we do not want to
arbitrarily change the number of methylated read counts. 'mC' carries
information linked to the amount of energy expended in the tissue associated
with concrete methylation changes. However, 'uC' is not linked to any energy
expended by the methylation machinery in the cells. In the bidimensional space
$P_i = (p_i, 1-p_i)$, reference *Ref0* corresponds to the point $P_i = (1, 0)$
at each cytosine site $i$, i.e., the value of methylation level at every
cytosine site in reference *Ref0* is 1. The analyses with respect to both
individual references, *Ref* and *Ref0*, will be performed in the downstream
steps.

# Hellinger divergence estimation
To perform the comparison between the uncertainty of methylation levels
from each group of individuals, control $(c)$ and treatment $(t)$, the
divergence between the  methylation levels of each individual is estimated with
respect to the same reference on the metric space formed by the vector set 
$P_i = (p_i, 1-p_i)$ and the Hellinger divergence $H$. Basically, the 
information divergence between the methylation levels of an individual $j$ and
reference sample $r$ is estimated according to the Hellinger divergence given 
by the  formula: 
$$H(\hat p_{ij},\hat p_{ir}) = w_i[(\sqrt{\hat p_{ij}} - \sqrt{\hat
p_{ir}})^2+(\sqrt{1-\hat p_{ij}} - \sqrt{1-\hat p_{ir}})^2]$$ where 
$w_i = 2 \frac{m_{ij} m_{ir}}{m_{ij} + m_{ir}}$, 
$m_{ij} = {n_i}^{mC_j}+{n_i}^{uC_j}+1$, $m_{ir} = {n_i}^{mC_r}+{n_i}^{uC_r}+1$ 
and $j \in {\{c,t}\}$. This equation for Hellinger divergence is given in
reference [@Basu2010], but other information theoretical divergences can be
used as well. Next, the information divergence for control (Breast_normal) and
treatment (Breast_cancer and Breast_metastasis) samples are estimated with
respect to the reference virtual individual. A Bayesian correction of counts
can be selected or not. In a Bayesian framework, methylated read counts are
modeled by a beta-binomial distribution, which accounts for both the
biological and sampling variations [@Hebestreit2013; @Robinson2014;
@Dolzhenko2014]. In our case we adopted the Bayesian approach suggested in
reference [@Baldi2001] (Chapter 3). In a Bayesian framework with uniform
priors, the methylation level can be defined as: $p = (mC + 1)/(mC + uC + 2)$. 

However, the most natural statistical model for replicated BS-seq DNA
methylation measurements is beta-binomial (the beta distribution is a prior
conjugate of binomial distribution). We consider the parameter $p$ (methylation
level) in the binomial distribution as randomly drawn from a beta distribution.
The hyper-parameters $\alpha$ and $\beta$ from the beta-binomial distribution
are interpreted as pseudo-counts. The information divergence is estimated here
using the function [estimateDivergence](https://genomaths.github.io/MethylIT_HTML_Manual/estimateDivergence.html):

```{r HD, include = TRUE}
HD = estimateDivergence(ref = Ref, indiv = LR, Bayesian = TRUE,  
                        min.coverage = 5, high.coverage = 300, 
                        percentile = 0.999, num.cores = 6L, tasks = 0L,
                        verbose = FALSE)
HD0 = estimateDivergence(ref = Ref0, indiv = LR, Bayesian = TRUE,  
                         min.coverage = 9, high.coverage = 300, 
                         percentile = 0.999, num.cores = 6L, tasks = 0L,
                         verbose = FALSE)
HD$Breast_cancer

```

Function 'estimateDivergence' returns a list of GRanges objects with the four 
columns of counts, the information divergence, and additional columns: 

  1. The original matrix of methylated ($c_i$) and unmethylated ($t_i$) read
     counts from control ($i=1$) and treatment ($i=2$) samples.
  2. "p1" and "p2": methylation levels for control and treatment, respectively.
  3. "bay.TV": total variation TV = p2 - p1.
  4. "TV": total variation based on simple counts: $TV=c1/(c1+t1)-c2/(c2+t2)$.
  5. "hdiv": Hellinger divergence.

If Bayesian = TRUE, results are based on the posterior estimations of
methylation levels $p1$ and $p2$. Filtering by coverage is provided at this
step which would be used unless previous filtering by coverage had been
applied. This is a pairwise filtering. Cytosine sites with 'coverage' >
'min.coverage' and 'coverage' < 'percentile' (e.g., 99.9 coverage percentile)
in at least one of the samples are preserved. The coverage percentile used is
the maximum estimated from both samples: reference and individual.

For some GEO datasets only the methylation levels for each cytosine site are
provided. In this case, Hellinger divergence can be estimated as given in
reference [@Sanchez2016]:
$$H(\hat p_{ij},\hat p_{ir}) = 2 \Big[(\sqrt{\hat p_{ij}} - \sqrt{\hat
p_{ir}})^2+(\sqrt{1-\hat p_{ij}} - \sqrt{1-\hat p_{ir}})^2\Big]$$

## _Histogram and boxplots of divergences estimated in each sample_
First, the data of interest (Hellinger divergences, "hdiv") are selected from
the GRanges objects:
```{r data4graphic, include = TRUE}
normal = HD$Breast_normal[, "hdiv"]
normal = normal[ normal$hdiv > 0.01 ]
metastasis = HD$Breast_metastasis[, "hdiv"]
metastasis = metastasis[ metastasis$hdiv > 0.01 ]
cancer = HD$Breast_cancer[, "hdiv"]
cancer = cancer[ cancer$hdiv > 0.01 ]

```
Next, a single GRanges object is built from the above set of GRanges objects
using the function [uniqueGRanges](https://genomaths.github.io/MethylIT_HTML_Manual/uniqueGRanges.html). 
Notice that the number of cores to use for parallel computation can be specified.
```{r hd4graphic, include = TRUE}
hd = uniqueGRanges(list(normal, cancer, metastasis), missing = NA,
                   verbose = FALSE, num.cores = 12L)
hd
```

Now, the Hellinger divergences estimated for each sample are in a single matrix
on the metacolumn of the GRanges object and we can proceed to build the
histogram and boxplot graphics for these data.

Finally, the data included in the graphic are:
```{r fig_data}
# Define an auxiliar function
fun_length <- function(x){
  return(data.frame(y = median(x) + 4, label = paste0("n = ", length(x))))
}

data <- data.frame(normal = hd$hdiv, cancer = hd$hdiv.1,
                   metastasis = hd$hdiv.2)
data = suppressMessages(melt(data))
colnames(data) <- c("Breast.tissue", "HD")
data = data[data$HD > 0, ]
DataFrame(data)
```


```{r fig1, fig.height = 7, fig.width = 20, message=FALSE}

p1 = ggplot(data, aes(x = HD, fill = Breast.tissue, colour = Breast.tissue)) + 
  geom_histogram(alpha = 0.5, bins = 50, position = "identity", na.rm = TRUE,
                 size = 0.7) + 
  theme(axis.title.x = element_text(face = "bold", size = 20),
        axis.text.x = element_text(face = "bold", size = 20, color = "black",
                                   hjust = 0.5, vjust = 0.75),
        axis.text.y = element_text(face = "bold", size = 20, color = "black"),
        axis.title.y = element_text(face = "bold", size = 20,color = "black"),
        legend.text = element_text(size = 20, face = "bold"),
        legend.title = element_text(size = 20, face = "bold")
        ) +
  ylab( "Counts" ) + 
  ggtitle("Histogram of Hellinger Divergence")

# For visualization purposes HD is limited to the interval 0 to 4
dt = data[ which(data$HD < 50), ] 
p2 = ggplot(dt,aes(x = Breast.tissue, y = HD , fill = Breast.tissue)) + 
  geom_boxplot(na.rm = TRUE) + 
  stat_summary(fun.data = fun_length, geom = "text",
               position = position_dodge(width = 0.9), vjust = 1,
               size = 6, fontface = "bold") +
  theme(axis.title.x = element_text(face = "bold", size = 20),
        axis.text.x = element_text(face = "bold", size = 20, color = "black",
                                   hjust = 0.5, vjust = 0.75),
        axis.text.y = element_text(face = "bold", size = 20, color = "black"),
        axis.title.y = element_text(face = "bold", size = 20,color = "black"),
        legend.position = "none"
        ) +
  ggtitle("Boxplot of Hellinger Divergence")
grid.arrange(p1, p2, ncol = 2)


```

Except for the tail, most of the methylation changes occurred under the area
covered by the density curve corresponding to the normal breast tissue. This is
theoretically expected. This area is explainable in statistical physical terms
and, theoretically, it should fit a Weibull distribution. The tails regions
cover the methylation changes that, with high probability, are not induced by
thermal fluctuation and are not addressed to stabilize the DNA molecule. These
changes are methylation signal. Professor David J. Miller (Department of
Electrical Engineering, Penn State) proposed modeling the distribution as a
mixed Weibull distribution to simultaneously describe the background
methylation noise and the methylation signal (personal communication, January,
2018). This model approach seems to be supported by the above histogram, but it
must be studied before being incorporated in a future version of Methyl-IT.

# Nonlinear fit of Weibull distribution
A basic requirement for the application of signal detection is the knowledge of
the probability distribution of the background noise. Probability distribution,
as a Weibull distribution model, can be deduced on a statistical
mechanical/thermodynamics basis for DNA methylation induced by thermal
fluctuations [@Sanchez2016]. Assuming that this background methylation
variation is consistent with a Poisson process, it can be distinguished from
variation associated with methylation regulatory machinery, which is
non-independent for all genomic regions [@Sanchez2016]. An
information-theoretic divergence to express the variation in methylation
induced by background thermal fluctuations will follow a Weibull distribution
model, provided that it is proportional to the minimum energy dissipated per
bit of information associated with the methylation change. The nonlinear fit to
a Weibull distribution model is performed by the function
[nonlinearFitDist](https://genomaths.github.io/MethylIT_HTML_Manual/nonlinearFitDist.html).

```{r nlms}
nlms = nonlinearFitDist(HD, column = 9, num.cores = 3L, verbose = FALSE)
nlms0 = nonlinearFitDist(HD0, column = 9, num.cores = 3L, verbose = FALSE)

nlms # this returns:

```
Cross-validations for the nonlinear regressions (R.Cross.val) were performed as
described in reference [@Stevens2009]. In addition, Stein's formula for
adjusted R squared ($\rho$) was used as an estimator of the average
cross-validation predictive power [@Stevens2009].

The goodness-of-fit of Weibull to the HD0 (*Ref0*) data is better than to HD
(*Ref*):
```{r nlm0}
nlms0

```
The goodness-of-fit indicators suggest that the fit to Weibull distribution
model for *Ref0* is better than for *Ref*.

# Signal detection
The information thermodynamics-based approach is postulated to provide greater
sensitivity for resolving true signal from the thermodynamic background within
the methylome [@Sanchez2016]. Because the biological signal created within the
dynamic methylome environment characteristic of plants is not free from
background noise, the approach, designated Methyl-IT, includes the application
of signal detection theory [@Greiner2000; @Carter2016; @Harpaz2013;
@Kruspe2017]. Signal detection is a critical step to increase sensitivity and
resolution of methylation signal by reducing the signal-to-noise ratio and
objectively controlling the false positive rate and prediction accuracy/risk.

## _Potential methylation signal_
The first estimation in our signal detection step is the identification of the
cytosine sites carrying potential methylation signal $PS$. The methylation
regulatory signal does not hold Weibull distribution and, consequently, for a
given level of significance $\alpha$ (Type I error probability, e.g. 
$\alpha = 0.05$), cytosine positions $k$ with information divergence 
$H_k >= H_{\alpha = 0.05}$ can be selected as sites carrying potential signals
$PS$. The value of $\alpha$ can be specified. For example, potential signals
with $H_k > H_{\alpha = 0.01}$ can be selected. For each sample, cytosine sites
are selected based on the corresponding fitted Weibull distribution model
estimated in the previous step. Additionally, since cytosine with 
$|TV_{d_k}| < 0.1$ are the most abundant sites, depending on the sample 
(experiment), cytosine positions $k$ with $H_k >= H_{\alpha = 0.05}$ and 
$|TV_{d_k}| < 0.1$ can be observed. To prevent the last situation we can select the
$PS$ with the additional constraint $|TV_{d_k}| > TV_0$, where $TV_0$ ('tv.cut') is
a user specified value. The $PS$ is detected with the function 
[getPotentialDIMP](https://genomaths.github.io/MethylIT_HTML_Manual/getPotentialDIMP.html):

```{r getPotentialDIMP}
PS = getPotentialDIMP(LR = HD, nlms = nlms, div.col = 9, alpha = 0.05,
                      tv.col = 7, tv.cut = 0.4)
PS0 = getPotentialDIMP(LR = HD0, nlms = nlms0, div.col = 9, alpha = 0.05,
                       tv.col = 7, tv.cut = 0.4)

PS$Breast_cancer

```

Notice that the total variation distance $TV_d=|TV|$ is an information
divergence as well and it can be used in place of Hellinger divergence
[@Sanchez2016]. The set of vectors $P_i = (p_i, 1-p_i)$ and distance function
$TV_d$ integrate a metric space. In particular, after takes the Manhattan
distance between healthy individuals and patients methylation levels we have:
$$M_d(p_{ij}, p_{ir})=\frac{1}{2}(|\hat p_{ij} - \hat p_{ir}| + |(1-\hat p_{ij}) - (1-\hat
p_{ir})|)=|\hat p_{ij} - \hat p_{ir}|=TV_d(p_{ij}, p_{ir})$$. 

That is, the absolute difference of methylation levels is a particular case of
Manhattan distance, which is named total variation distance $TV_d$. The
quantitative effect of the vector components $1-\hat p_{ij}$ and $1-\hat p_{ir}$
(in our case, the effect of unmethylated read counts) is not present in $TV_d$ as
in $H(\hat p_{ij},\hat p_{ir})$.

## _Histogram and boxplots of methylation potential signals_
As before, a single GRanges object is built from the above set GRanges objects
using the function [uniqueGRanges](https://genomaths.github.io/MethylIT_HTML_Manual/uniqueGRanges.html),
and the Hellinger divergences of the cytosine sites carrying $PS$ (for each
sample) are located in a single matrix on the metacolumn of the GRanges object.

```{r fig2, fig.height = 7, fig.width = 20}
ps = uniqueGRanges(PS, missing = NA, verbose = FALSE, num.cores = 12L)
dat = data.frame(normal = ps$hdiv, cancer = ps$hdiv.1, metastasis = ps$hdiv.2)
dat = suppressMessages(melt(dat))
colnames(dat) <- c("Breast.tissue", "HD")
idx <- which(is.na(dat$HD))
dat <- dat[-idx, ] # To remove missing data
DataFrame(dat)

p1 = ggplot(dat, aes(x = HD, fill = Breast.tissue, colour = Breast.tissue)) + 
  geom_histogram(alpha = 0.5, bins = 50, position = "identity", na.rm = TRUE,
                 size = 0.7)  + ylab("Counts") + xlim(7,50) +
theme(axis.title.x = element_text(face = "bold", size = 20),
        axis.text.x = element_text(face = "bold", size = 20, color = "black",
                                   hjust = 0.5, vjust = 0.75),
        axis.text.y = element_text(face = "bold", size = 20, color = "black"),
        axis.title.y = element_text(face = "bold", size = 20,color = "black"),
        legend.text = element_text(size = 20, face = "bold"),
        legend.title = element_text(size = 20, face = "bold")
        ) +
    ggtitle("Histogram for Potential methylation signal")

p2 = ggplot(dat,aes(x = Breast.tissue, y = HD , fill = Breast.tissue)) +
  geom_boxplot(na.rm = TRUE) +
  stat_summary(fun.data = fun_length, geom = "text", na.rm = TRUE, 
               position = position_dodge(width = 0.9), vjust = 1,
               size = 6, fontface = "bold") +
  theme(axis.title.x = element_text(face = "bold", size = 20),
        axis.text.x = element_text(face = "bold", size = 20, color = "black",
                                   hjust = 0.5, vjust = 0.75),
        axis.text.y = element_text(face = "bold", size = 20, color = "black"),
        axis.title.y = element_text(face = "bold", size = 20,color = "black"),
        legend.position = "none"
        ) +
      ggtitle("Boxplot for Potential methylation signal")

grid.arrange(p1, p2, ncol = 2)


```

# Cutpoint estimation
Laws of statistical physics can account for background methylation, a response
to thermal fluctuations that presumably functions in DNA stability
[@Sanchez2016]. True signal is detected based on the optimal cutpoint
[@lopez2014], which can be estimated from the area under the curve (AUC) of a
receiver operating characteristic (ROC) curve built from a logistic regression
performed with the potential signals from controls and treatments. The ROC AUC
is equivalent to the probability that a randomly-chosen positive instance is
ranked more highly than a randomly-chosen negative instance [@Fawcett2005]. In
the current context, the AUC is equivalent to the probability to distinguish a
randomly-chosen methylation regulatory signal induced by the treatment from a
randomly-chosen signal in the control.

The need for the application of (what is now known as) signal detection in
cancer research was pointed out by Youden in the midst of the last century
[@Youden1950]. In the next example, the simple cutpoint estimation available in
Methyl-IT is based on the aplication of [Youden
index](https://en.wikipedia.org/wiki/Youden%27s_J_statistic) [@Youden1950].
Although cutpoints are estimated for a single variable, the classification
performance can be evaluated for several variables and appliying different model
classifiers. A optimal cutpoint distinguishs disease stages from healthy
indiviadual. The performance of this classification is given in the ouput of
function
[estimateCutPoint](https://genomaths.github.io/MethylIT_HTML_Manual/estimateCutPoint.html).
A model classifier can be requested for further predicitons and its
classification performance is also provided. Below, the selected model
classifier is a quadratic discriminant analysis (QDA) (*classifier1* = "qda",
*clas.perf* = TRUE). Four predictor variables are available: the Hellinger
divergence of methylation levels (*hdiv*), total variation distance (*TV*,
absolute difference of methylation levels), relative position of cytosine site
in the chromosome (*pos*), and the logarithm base two of the probability to
obseve a Hellinger divergence value $H$ greater than the critical value
$H_{\alpha = 0.05}$ (values given as probabilities in object PS, *wprob*). 

Notice that the cutpoint can be estimated for any of the two currently available
information divergences: *Hellinger divergence* (*div.col* = 9) or the *total*
*variation distance* (*div.col* = 7).

```{r cutpoints}
cutpoint = estimateCutPoint(LR = PS, control.names = "Breast_normal", 
                             treatment.names = c("Breast_cancer", 
                                                 "Breast_metastasis"),
                             classifier1 = "qda", simple = TRUE,
                             column = c(hdiv = TRUE, TV = TRUE, 
                                        wprob = TRUE, pos = TRUE),
                             div.col = 7, clas.perf = TRUE)
cutpoint

```

In the current case, the classification based on *Youden index* and based only 
on $TV_d$ is not good. The specificity is 0. That is, this model fails in to 
properly distinguish healthy individuals from patients:  
```{r youden}
cutpoint$initModelConfMatrix
```
However, the request for a QDA model classifier solved the problem. Still the 
cutpoint based on Youden index is used, but QDA model include more information 
about each cytosine site, which improve the classification performance. Below, 
model confusion matrix is requested, which include the whole set of DMPs. The 
model performance in the test set is given at *cutpoint$testSetPerformance*.

```{r youden2}
cutpoint$modelConfMatrix

```
## _Searching the optimal cutpoint with model classifiers_

The previous result can be improved requesting a cutpoint search based on QDA
model classifier:

```{r cutpoints2}
cutpoint = estimateCutPoint(LR = PS, control.names = "Breast_normal", 
                             treatment.names = c("Breast_cancer", 
                                                 "Breast_metastasis"),
                             classifier1 = "qda", simple = FALSE,
                             column = c(hdiv = TRUE, TV = TRUE, 
                                        wprob = TRUE, pos = TRUE),
                             find.cut = TRUE, cut.interval = c(0.5, 0.8), 
                             cut.incr = 0.01, stat = 0, num.cores = 6,
                             div.col = 9, clas.perf = TRUE)
cutpoint
```
The cutpoint $H=6.682$ permitted the detection of more DMPs in both groups: 
control and patients.

```{r findcut}
cutpoint$modelConfMatrix
```
Similar analysis is applied to the alternative datset PS0:
```{r cutpoints0}
cutpoint0 = estimateCutPoint(PS0, control.names = "Breast_normal",
                              treatment.names = c("Breast_cancer", 
                                                   "Breast_metastasis"),
                              classifier1 = "qda", simple = FALSE,
                              find.cut = TRUE, cut.interval = c(0.5, 0.8), 
                              cut.incr = 0.01, stat = 0, num.cores = 6,
                              div.col = 9, verbose = FALSE)
cutpoint0

```


```{r}
cutpoint0$modelConfMatrix
```

# DMPs
Cytosine sites carrying a methylation signal are designated _differentially
informative methylated positions_ (DMPs). The probability that a DMP is not
induced by the treatment is given by the probability of false alarm ($P_{FA}$,
false positive). That is, the biological signal is naturally present in the
control as well as in the treatment. Each DMP is a cytosine position carrying
a significant methylation signal, which may or may not be represented within a
differentially methylated position (DMP) according to Fisher's exact test (or
other current tests). A DMP is a DNA cytosine position with high probability
to be differentially methylated or unmethylated in the treatment with respect
to a given control. Notice that the definition of DMP is not deterministic in
an ordinary sense, but stochastic-deterministic in physico-mathematical terms.

DMPs are selected with the function [selectDIMP](https://genomaths.github.io/MethylIT_HTML_Manual/selectDIMP.html)
```{r DMPs}
DMPs = selectDIMP(PS, div.col = 9, cutpoint = 6.682	)
DMPs0 = selectDIMP(PS0, div.col = 9, cutpoint = 1.1675)

```

## _Histogram and boxplots of DMPs_
The cutpoint detected with the signal detection step is very close (in this
case) to the Hellinger divergence value $H_{\alpha = 0.05}$ estimated for
cancer tissue. The natural methylation regulatory signal is still present in a
patient with cancer and reduced during the metastasis step. This signal is
detected here as a false alarm ($P_{FA}$, false positive)

The list of GRanges with DMPs are integrated into a single GRanges object with 
the matrix of 'hdiv' values on its metacolumn:   
```{r DMPs4graphics}
dimp = uniqueGRanges(DMPs, missing = NA, verbose = FALSE, num.cores = 12L)
dat <- data.frame(normal = dimp$hdiv, cancer = dimp$hdiv.1,
                  metastasis = dimp$hdiv.2)
dat = suppressMessages(melt(dat))
colnames(dat) <- c("Breast.tissue", "HD")
dt = dat[ which(dat$HD < 50), ]
cutp = 6.682	
DataFrame(dt)

```

The multiplot with the histogram and the boxplot can now built:
```{r fig3, fig.height = 7, fig.width = 20}
p1 = ggplot(data, aes(x = HD, fill = Breast.tissue, colour = Breast.tissue)) +
  geom_histogram(alpha = 0.5, bins = 50, position = "identity", na.rm = TRUE,
                 size = 0.7) + ylab( "Counts" ) +
  geom_vline(xintercept = cutp, color = "red", linetype = "dashed") +
  annotate(geom = "text", x = cutp + 5, y = -11000, fontface = 2, size = 6,
           label = paste0("cutopoint = ", cutp), family = "serif") +
  annotate(geom = "text", x = cutp + 5, y = 4.1e5, label = "DMPs",
           fontface = 2, size = 6, family = "serif") +
  geom_segment(aes(x = cutp, xend = 60, y = 4e5, yend = 4e5),
               arrow = arrow(length = unit(0.5, "cm")), color = "blue") +
  theme(axis.title.x = element_text(face = "bold", size = 20, family = "serif"),
        axis.text.x = element_text(face = "bold", size = 20, color = "black",
                                   hjust = 0.5, vjust = 0.75, family = "serif"),
        axis.text.y = element_text(face = "bold", size = 20, color = "black",
                                   family = "serif"),
        axis.title.y = element_text(face = "bold", size = 20,color = "black",
                                    family = "serif"),
        legend.text = element_text(size = 20, face = "bold", family = "serif"),
        legend.title = element_text(size = 20, face = "bold", family = "serif")
  )

p2 = ggplot(dt,aes(x = Breast.tissue, y = HD , fill = Breast.tissue)) +
  geom_boxplot(na.rm = TRUE) + ylab("DMPs HD values") +
  stat_summary(fun.data = fun_length, geom = "text",
               position = position_dodge(width = 0.9), vjust = 2,
               size = 6, fontface = "bold", family = "serif") +
  xlab("Breast tissue") +
  theme(axis.title.x = element_text(face = "bold", size = 20),
        axis.text.x = element_text(face = "bold", size = 20, color = "black",
                                   hjust = 0.5, vjust = 0.75, family = "serif"),
        axis.text.y = element_text(face = "bold", size = 20, color = "black",
                                   family = "serif"),
        axis.title.y = element_text(face = "bold", size = 20,color = "black",
                                    family = "serif"),
        legend.position = "none"
  )
grid.arrange(p1, p2, ncol = 2)

```

## _Venn Diagram of DMPs_
The Venn diagram of DMPs reveals that the number cytosine site carrying
methylation signal with a divergence level comparable to that observed in
breast tissues with cancer and metastasis is relatively small (2797 DMPs). The
number of DMPs decreased in the breast tissue with metastasis, but, as shown
in the last boxplot, the intensity of the signal increased.

```{r fig4, fig.height = 5, fig.width = 7}
n12 = length(GenomicRanges::intersect(DMPs$Breast_normal,
                                      DMPs$Breast_cancer))
n13 = length(GenomicRanges::intersect(DMPs$Breast_normal,
                                      DMPs$Breast_metastasis))
n23 = length(GenomicRanges::intersect(DMPs$Breast_cancer,
                                      DMPs$Breast_metastasis))
n123 = length(Reduce(GenomicRanges::intersect, 
                     list(DMPs$Breast_normal, DMPs$Breast_cancer,
                          DMPs$Breast_metastasis)))

grid.newpage()
v = draw.triple.venn(area1 = length(DMPs$Breast_normal),
                     area2 = length(DMPs$Breast_cancer),
                     area3 = length(DMPs$Breast_metastasis),
                     n12 = n12, n23 = n23, n13 = n13, n123 = n123,
                     category = c("Breast_normal", "Breast_cancer", 
                                  "Breast_metastasis"),
                     lty = rep("blank", 3), fill = c("blue", "yellow",
                                                     "magenta"),
                     alpha = c(0.1, 0.2, 0.3), 
                     cat.pos = c(-80, 90, 0),
                     cat.col = c("blue", "darkorange", "red"),
                     cat.dist = c( -0.1, -0.08, -0.26),
                     cex = rep(1.7, 7),
                     cat.cex = c( 1.5, 1.5, 1.5),
                     label.col = c( "blue", "darkorange", "darkorange",
                                    "red", 
                                    "white", "red", "red"),
                     scaled = TRUE)
grid.draw(v)


```


Notice that natural methylation regulatory signals (not induced by the
treatment) are present in both groups, control and treatment. The signal
detection step permits us to discriminate the "ordinary" signals observed in
the control from those induced by the treatment (a disease in the current
case). In addtion, this diagram reflects a classification of DMPs only based
on the cytosine positions. That is, this Venn diagram cannot tell us whether
DMPs at the same postion can be distinguishable or not. For example, DMPs at
the same positions in control and treatment can happend with different
probabilities estimated from their corresponding fitted Weibull distributions
(see below).

## _Venn Diagram of DMPs for reference Ref0_

```{r fig5, fig.height = 5, fig.width = 7}
DMPs0 = selectDIMP(PS0, div.col = 9, cutpoint = 3.514418)

n12 = length(GenomicRanges::intersect(DMPs0$Breast_normal,
                                      DMPs0$Breast_cancer))
n13 = length(GenomicRanges::intersect(DMPs0$Breast_normal,
                                      DMPs0$Breast_metastasis))
n23 = length(GenomicRanges::intersect(DMPs0$Breast_cancer,
                                      DMPs0$Breast_metastasis))
n123 = length(Reduce(GenomicRanges::intersect, 
                     list(DMPs0$Breast_normal, DMPs0$Breast_cancer,
                          DMPs0$Breast_metastasis)))

grid.newpage()
v = draw.triple.venn(area1 = length(DMPs0$Breast_normal),
                     area2 = length(DMPs0$Breast_cancer),
                     area3 = length(DMPs0$Breast_metastasis),
                     n12 = n12, n23 = n23, n13 = n13, n123 = n123,
                     category = c("Breast_normal", "Breast_cancer", 
                                  "Breast_metastasis"),
                     lty = rep("blank", 3), fill = c("blue", "yellow",
                                                     "magenta"),
                     alpha = c(0.1, 0.2, 0.3), 
                     cat.pos = c(-80, 90, 0),
                     cat.col = c("blue", "darkorange", "red"),
                     cat.dist = c( -0.1, -0.08, -0.26),
                     cex = rep(1.7, 7),
                     cat.cex = c( 1.5, 1.5, 1.5),
                     label.col = c( "blue", "darkorange", "darkorange",
                                    "red", 
                                    "white", "red", "red"),
                     scaled = TRUE)
grid.draw(v)


```


# Differentially informative methylated genomic regions (DIMRs)
Our degree of confidence in whether DMP counts in both groups of samples,
control and treatment, represent true biological signal was determined in the
signal detection step. To estimate DIMRs, we followed similar steps to those
proposed in Bioconductor R package DESeq2 [@Love2014], but our GLM test looks
for statistical difference between the groups based on gene-body DMP counts
overlapping a given genomic region rather than read counts. The regression
analysis of the generalized linear model (GLM) with logarithmic link was
applied to test the difference between group counts. The fitting algorithmic
approaches provided by 'glm' and 'glm.nb' functions from the R packages stat
and MASS, respectively, were used for Poisson (PR), Quasi-Poisson (QPR) and
Negative Binomial (NBR) linear regression analyses, respectively.

## _Differentially methylated genes (DMGs)_
We shall call DMGs those DIMRs restricted to gene-body regions. DMGs are
detected using function 'countTest'. We used computational steps from DESeq2
packages. In the current case we follow the steps:
```{r AG}
# To load human gene annotation 
AG = import(con = paste0("ftp://ftp.ensembl.org/pub/release-91/gff3/",
                         "homo_sapiens/Homo_sapiens.GRCh38.91.gff3.gz"))
genes = AG[ AG$type == "gene", c( "gene_id", "biotype" ) ]
genes = genes[ genes$biotype == "protein_coding", "gene_id" ]
seqlevels(genes, "coarse") <- "13" # To keep a consistent chromosome annotation
seqlevels(genes) <- "chr13"

```

Function [getDIMPatGenes](https://genomaths.github.io/MethylIT_HTML_Manual/getDIMPatGenes.html)
is used to count the number of DMPs at gene-body.
The operation of this function is based on the '*findOverlaps*' function from the
'*GenomicRanges*' Bioconductor R package. The '*findOverlaps*' function has several
critical parameters like, for example, '*maxgap*', '*minoverlap*', and
'*ignore.strand*'. In our function [getDIMPatGenes](https://genomaths.github.io/MethylIT_HTML_Manual/getDIMPatGenes.html), 
except for setting ignore.strand = TRUE and *type = "within"*, we preserve the rest of default
'findOverlaps' parameters. In this case, these are important parameter settings
because the local mechanical effect of methylation changes on a DNA region
where a gene is located is independent of the strand where the gene is encoded.
That is, methylation changes located in any of the two DNA strands inside the
gene-body region will affect the flexibility of the DNA molecule [@Choy2010;
@Severin2011].

```{r getDIMPatGenes}
DMPsBN = getDIMPatGenes(GR = DMPs$Breast_normal, GENES = genes)
DMPsBC = getDIMPatGenes(GR = DMPs$Breast_cancer, GENES = genes)
DMPsBM = getDIMPatGenes(GR = DMPs$Breast_metastasis, GENES = genes)

```

The number of DMPs on the strand where a gene is encoded is obtained by
setting ignore.strand = FALSE. However, for the current example results will be
the same since the datasets downloaded from GEO do not have strand information.
Next, the above GRanges objects carrying the DMP counts from each sample are
grouped into a single GRanges object. Since we have only one control, to
perform group comparison and to move forward with this example, we duplicated
'Breast_normal' sample. Obviously, the confidence on the results increases with
the number of sample replications per group (in this case, it is only an
illustrative example on how to perform the analysis, since a fair comparison
requires for more than one replicate in the control group).

```{r uniqueGRanges}
Genes.DMPs = uniqueGRanges( list(DMPsBN[, 2], DMPsBN[, 2],
                                  DMPsBC[, 2], DMPsBM[, 2]),
                             type = "equal", verbose = FALSE,
                             ignore.strand = TRUE )
colnames( mcols(Genes.DMPs)) <- c("Breast_normal", "Breast_normal1",
                                   "Breast_cancer", "Breast_metastasis")

```

Next, the set of mapped genes are annotated 
```{r subsetByOverlaps}
GeneID = subsetByOverlaps(genes, Genes.DMPs, type = "equal", 
                          ignore.strand = FALSE)
names( Genes.DMPs ) <- GeneID$gene_id
Genes.DMPs
```

Now, we build a 'DESeqDataSet' object using functions DESeq2 package.  
```{r glmDataSet}
## An experiment design is set.
colData <- data.frame(condition = factor(c("BN","BN","BC","BC")),
                       c("Breast_normal","Breast_normal1",
                         "Breast_cancer","Breast_metastasis"),
                       row.names = 2)
## A RangedGlmDataSet is created
ds <- glmDataSet(GR = Genes.DMPs, colData = colData)
ds
```

DMG analysis is performed with the function [countTest](https://genomaths.github.io/MethylIT_HTML_Manual/countTest2.html)
```{r countTest}
DMGs = countTest2(DS = ds, num.cores = 4L, minCountPerIndv = 8, 
                  countFilter = TRUE, maxGrpCV = c(1, 0.5),
                 Minlog2FC = 1, pvalCutOff = 0.05,
                 MVrate = .95, verbose = FALSE)
DMGs

```


## _DMGs for reference Ref0_

```{r DIMR0}
DMPs0BN = getDIMPatGenes(GR = DMPs0$Breast_normal, GENES = genes)
DMPs0BC = getDIMPatGenes(GR = DMPs0$Breast_cancer, GENES = genes)
DMPs0BM = getDIMPatGenes(GR = DMPs0$Breast_metastasis, GENES = genes)

Genes.DMPs0 = uniqueGRanges( list(DMPs0BN[, 2], DMPs0BN[, 2],
                                   DMPs0BC[, 2], DMPs0BM[, 2]),
                              type = "equal", verbose = FALSE,
                              ignore.strand = TRUE )
colnames( mcols(Genes.DMPs0)) <- c("Breast_normal", "Breast_normal1",
                                    "Breast_cancer", "Breast_metastasis")

GeneID = subsetByOverlaps(genes, Genes.DMPs0, type = "equal", 
                          ignore.strand = FALSE)
names( Genes.DMPs0 ) <- GeneID$gene_id

## An experiment design is set.
colData <- data.frame(condition = factor(c("BN","BN","BC","BC")),
                       c("Breast_normal","Breast_normal1",
                         "Breast_cancer","Breast_metastasis"),
                       row.names = 2)
## A RangedGlmDataSet is created
ds0 <- glmDataSet(GR = Genes.DMPs0, colData = colData)
ds0

DMGs0 = countTest2(DS = ds0, num.cores = 4L, minCountPerIndv = 8, 
                  countFilter = TRUE, maxGrpCV = c(1, 0.5),
                 Minlog2FC = 1, pvalCutOff = 0.05,
                 MVrate = .95, verbose = FALSE)

DMGs0
```


_**BRCA2, a breast cancer associated risk gene, is found between the DMG**s_

```{r DMGs0}
# DMGs0
DMGs0[ grep( "ENSG00000139618", names(DMGs0) ) ]

```

# Monte Carlo Evaluation of the Model Classification Performance
The regulatory methylation signal is an output from a natural process that
continuously takes place across the ontogenetic development of the organism.
Therefore, we expect to see methylation signal in natural, ordinary conditions.
Function 'evaluateDIMPclass' can be used to perform a classification of DMPs
into two classes: DMPS from control and DMPs from treatment samples, as well
as an evaluation of the classification performance (for more details see
[evaluateDIMPclass](https://genomaths.github.io/MethylIT_HTML_Manual/evaluateDIMPclass.html)). 
As a matter of fact this function is called by function [estimateCutPoint](https://genomaths.github.io/MethylIT_HTML_Manual/estimateCutPoint.html).
Additional feature not used before is the possibility to perform Monte Carlo evaluation 
of the model classifier performance. In above split of the sample into two subsets, 
training and test datasets, we would be just lucky getting the right proportion 
that yields the best classification performance. What about is the random split is
repeat 300 times?

The performance of the logistic classifier using reference 'Ref' is:
```{r mc1}
performance <- evaluateDIMPclass(LR = DMPs, control.names = "Breast_normal",
                              treatment.names = c("Breast_cancer", 
                                                  "Breast_metastasis"),
                              column = c(hdiv = TRUE, TV = TRUE, 
                                         wprob = TRUE, pos = TRUE),
                              classifier = "qda", prop = 0.6, num.boot = 300,
                              output = "mc.val", num.cores = 6L)
performance
```
```{r mc2}
performance0 <- evaluateDIMPclass(LR = DMPs0, control.names = "Breast_normal",
                              treatment.names = c("Breast_cancer", 
                                                  "Breast_metastasis"),
                              column = c(hdiv = TRUE, TV = TRUE, 
                                         wprob = TRUE, pos = TRUE),
                              classifier = "qda", prop = 0.6, num.boot = 300,
                              output = "mc.val", num.cores = 6L)
performance0
```

# Concluding remarks

Results fundamentaly depend on the use of the proper reference samples. If there is
not enough control samples to create an independent reference virtual individual, then
the centroid from the control group can be used instead, which can be built with function 
[poolFromGRlist](https://genomaths.github.io/MethylIT_HTML_Manual/poolFromGRlist.html).

The *reference* is used to put individuals from control and patients in the same
*system of reference*. This permits to evaluate how big is the spontaneous
natural variation in the control population. *It does not matter how statistically* 
*significant a DMP would be in the group of patients*. *What really matter is how big is the* 
*probality to observe the same methylation event in the control group in repect to the patient group*. 
Decisions are ultimately made on the basis of such classification probabilities at single 
cytosine positions. The signal detection based approach implemented in MethylIT
is addressed to confront the mentioned issue and, in consequence, the level of 
resolution of MethylIT approach reach the single cytosine site. 


# Acknowledgments
We thank Professor David J Miller for valuable conversations and suggestions on
our mathematical modeling.
# Funding The work was supported by funding from NSF-SBIR
(2015-33610-23428-UNL) and the Bill and Melinda Gates Foundation (OPP1088661).

## S2. Session Information

```{r, echo=FALSE}
sessionInfo()

```


# References. 

